# Biomarker discovery as causal effect estimation {#sec-biotmle}

```{r}
#| label: setup-biotmle
#| include: false
library(SuperLearner)
library(ggplot2)
source(here::here("practical/R/utils.R"))
source(here::here("practical/R/moderate.R"))
source(here::here("practical/R/learners.R"))
source(here::here("practical/R/sim_cpg.R"))
methy <- read.csv(here::here("practical/data/su2016_methylation.csv.gz"))
set.seed(11249)
```

::: {.callout-note title="In a nutshell"}
We rebuild the `biotmle` recipe from its parts. First, we estimate the
average treatment effect (ATE) of smoking on methylation at each of many CpG
sites by targeted maximum likelihood (TMLE) with machine learning. Then we
test all CpGs with variance moderation.

**Roadmap:** (1) the target parameter, and when a regression coefficient
equals it → (2) TMLE by hand, on a synthetic CpG with a known effect →
(3) misspecified GLMs, correct GLMs, and Super Learner → (4) the real data,
one CpG and then all of them → (5) variance moderation and multiple testing.
:::

## The data: smoking and DNA methylation

@su2016distinct measured DNA methylation in whole blood, at roughly 470,000
CpG sites, in 253 people: 172 who had ever smoked and 81 who never had. We
use their data to ask: **which CpG sites are affected by smoking?**

The data on one person are $O = (L, A, Y) \sim \P_0 \in \mathcal{P}$, where
$\P_0$ is the true, unknown data-generating distribution and $\mathcal{P}$
is the nonparametric statistical model; we observe $n$ i.i.d. copies
$O_1, \ldots, O_n$. Here,

- $L$ are baseline covariates: sex, age, race, and the estimated proportions
  of six leukocyte types (`CD8T`, `CD4T`, `NK`, `Bcell`, `Mono`, `Gran`).
  Blood cell composition differs by smoking status and strongly affects
  methylation, so it is a confounder;
- $A \in \{0, 1\}$ is smoking status (`smoker`: 1 = ever, 0 = never);
- $Y = (Y_b : b = 1, \ldots, B)$ are the methylation levels (M-values) at
  each CpG site $b$.

```{tikz}
%| label: fig-dag-biotmle
%| echo: false
%| fig-cap: "DAG for one CpG site $b$, in time order: covariates $L$ confound the effect of smoking $A$ on methylation $Y_b$."
%| out-width: 45%
\usetikzlibrary{arrows.meta}
\tikzset{>={Stealth[length=2.2mm]}, every node/.style={minimum size=6mm},
  arc/.style={->, thick}}
\begin{tikzpicture}[node distance=2cm]
  \node (L) {$L$};
  \node[right of=L] (A) {$A$};
  \node[right of=A] (Y) {$Y_b$};
  \draw[arc] (L) -- (A);
  \draw[arc] (A) -- (Y);
  % arcs leave and enter nodes at their tops, so arrowheads do not collide
  \draw[arc] (L.north) to[out=50, in=130] (Y.north);
\end{tikzpicture}
```

To keep computation fast, the file contains $B = 300$ CpG sites: 150 chosen at
random among those flagged by a `limma` screen in the full data (including
four well-known smoking-related CpGs, e.g., `cg05575921` in the *AHRR* gene),
and 150 chosen at random among those that were not.

```{r}
#| label: look-data
covars <- c("sex", "age", "race", "CD8T", "CD4T", "NK", "Bcell", "Mono", "Gran")
cpgs <- grep("^cg", names(methy), value = TRUE)
A <- methy$smoker
L <- methy[, covars]
n <- nrow(methy)

dim(methy)
table(smoker = A)
```

## The target parameter

Let $g_0(L) = \P_0(A = 1 \mid L)$ be the propensity score and
$\overline{Q}_{0,b}(A, L) = \E_0[Y_b \mid A, L]$ the outcome regression for
CpG $b$. For each CpG, our target is the ATE of smoking on methylation,
which (under no unmeasured confounding and positivity) is identified by
$$
\psi_{0,b} = \Psi_b(\P_0) = \E_0\big[ \overline{Q}_{0,b}(1, L) -
\overline{Q}_{0,b}(0, L) \big].
$$
This is a *variable importance measure*: it ranks CpG sites by how much
smoking would change their methylation, without committing to a model for
$\overline{Q}_{0,b}$. Throughout, a subscript $n$ denotes an estimate (e.g.,
$g_n$, $\overline{Q}_{n,b}$, $\psi_{n,b}$).

### Exercise 1.1 ✏️ {#sec-ex11}

::: {.callout-note title="When is a regression coefficient the ATE?"}
The usual approach (as in `limma`) fits, for each CpG, the working linear
model $\overline{Q}_{0,b}(a, l) = \beta_0 + \beta_1 a + \gamma^\top l$, and
tests $\beta_1 = 0$.

a. Suppose this linear model is correct. Show that $\psi_{0,b} = \beta_1$.
b. Now suppose the effect of smoking varies with age, so that the correct
   model is $\overline{Q}_{0,b}(a, l) = \beta_0 + \beta_1 a + \beta_2 a
   \cdot \text{age} + \gamma^\top l$. What is $\psi_{0,b}$ now? Is
   $\beta_1$ still the ATE?
c. In (b), what would $\beta_1$ in the *main-terms* model of (a) estimate?
   Is there a study design under which it would still estimate the ATE?
:::

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
a. $\overline{Q}_{0,b}(1, l) - \overline{Q}_{0,b}(0, l) = (\beta_0 + \beta_1
   + \gamma^\top l) - (\beta_0 + \gamma^\top l) = \beta_1$ for every $l$, so
   averaging over $L$ gives $\psi_{0,b} = \beta_1$.
b. Now $\overline{Q}_{0,b}(1, l) - \overline{Q}_{0,b}(0, l) = \beta_1 +
   \beta_2 \, \text{age}$, so $\psi_{0,b} = \beta_1 + \beta_2
   \E_0[\text{age}]$. The coefficient $\beta_1$ is the effect at age 0, which
   is rarely of scientific interest.
c. The main-terms coefficient converges to (roughly) a weighted average of
   the age-specific effects, with weights proportional to the conditional
   variance of smoking, $\V_0(A \mid L)$: people whose smoking status is
   hard to predict from $L$ count more. It is not $\psi_{0,b}$ in general.
   If $A$ were randomized ($A \indep L$), the main-terms coefficient would be
   consistent for the ATE even though the model is wrong. Smoking is not
   randomized, so here we need an estimator whose validity does not hinge on
   the form of $\overline{Q}_{0,b}$.

**Take-away:** $\psi_{0,b}$ is defined without reference to any model;
$\beta_1$ *is* $\psi_{0,b}$ only if the linear model is right (or $A$ is
randomized).
:::
:::

## Efficient estimators: the one-step estimator and TMLE

From the lecture, the efficient influence function (EIF) of $\psi_{0,b}$ is
$$
D_b(O; \P_0) = H_0(A, L) \big(Y_b - \overline{Q}_{0,b}(A, L)\big)
  + \overline{Q}_{0,b}(1, L) - \overline{Q}_{0,b}(0, L) - \psi_{0,b},
\qquad
H_0(A, L) = \frac{A}{g_0(L)} - \frac{1 - A}{1 - g_0(L)}.
$$
Given estimates $g_n$ and $\overline{Q}_{n,b}$, the *one-step* estimator
corrects the plug-in estimate by the average of the estimated EIF:
$$
\psi^+_{n,b} = \frac{1}{n} \sum_{i = 1}^n \big\{\overline{Q}_{n,b}(1, L_i) -
  \overline{Q}_{n,b}(0, L_i)\big\} + \frac{1}{n} \sum_{i = 1}^n H_n(A_i, L_i)
  \big(Y_{b,i} - \overline{Q}_{n,b}(A_i, L_i)\big).
$$
The correction removes the first-order bias of the plug-in, but the result
can fall outside the parameter's range.

**TMLE** removes the same bias differently: it *updates* the initial fit
$\overline{Q}_{n,b}$ to a targeted fit $\overline{Q}^\star_{n,b}$ chosen so
that the average of the estimated EIF is zero, and then plugs it in,
$$
\psi^\star_{n,b} = \frac{1}{n} \sum_{i = 1}^n \big\{\overline{Q}^\star_{n,b}(1,
L_i) - \overline{Q}^\star_{n,b}(0, L_i)\big\}.
$$
TMLE is a **substitution (plug-in) estimator**, $\psi^\star_{n,b} =
\Psi_b(\P^\star_n)$: it is the value of the target parameter at an estimate
of the data distribution, so it always respects the parameter's bounds.
Respecting the bounds also tends to make TMLE more stable in finite samples.
It shares the one-step estimator's large-sample properties: it is efficient,
its standard error is $\sqrt{\sigma^2_{n,b} / n}$ with $\sigma^2_{n,b}$ the
empirical variance of the estimated EIF, and it is *doubly robust*:
consistent if either $g_n$ or $\overline{Q}_{n,b}$ is consistent.

For a continuous outcome, the update is made on the unit scale, so that
the targeted fit respects the observed range of the outcome
[@gruber2010targeted]. The helper functions in `practical/R/utils.R` carry
out the parts of each step that are not about TMLE itself:

1. Scale $Y_b$ to $[0, 1]$, using its observed range.
   - *Helper:* `scale_to_unit()`.
2. Estimate $g_0$ and (on the unit scale) $\overline{Q}_{0,b}$.
   *Helpers:* `estimate_g()` and `estimate_Q()` fit the nuisance parameters
   with `SuperLearner()`, using the library of learners specified.
   `estimate_g()` passes its predictions through `bound_propensity()`, which
   truncates predicted probabilities away from 0 and 1, to avoid instability in
   $H_n$ (note: truncation does trade a bit of bias for improved variance).
   `bound_precision()` keeps the predicted outcomes strictly inside $(0, 1)$,
   so that their logits (step 4) are finite.
3. Compute the *clever covariate* $H_n(A, L)$, at the observed $A$ and at
   $A = 1$ and $A = 0$. *By hand.*
4. *Fluctuate*: fit the logistic regression $\operatorname{logit}
   \overline{Q}_{n,b}(\epsilon)(A, L) = \operatorname{logit}
   \overline{Q}_{n,b}(A, L) + \epsilon H_n(A, L)$, with the initial fit as an
   offset, giving $\epsilon_n$. *By hand*, with `glm()`.
5. Update, $\overline{Q}^\star_{n,b} = \overline{Q}_{n,b}(\epsilon_n)$, and
   plug in. Fitting $\epsilon$ by maximum likelihood makes the average of
   the EIF zero: TMLE *solves the EIF estimating equation*. *By hand.*
6. Compute the EIF at the targeted fit for the standard error, and scale
   back to the original units (multiplying by the range of $Y_b$, as
   `scale_from_unit()` does). *By hand.*

`tmle_ate()` wraps all six steps; we use it once we have done them by hand.

### A synthetic CpG with a known effect

Before turning to the real data, we try TMLE on a **synthetic CpG** for which
we know the true causal effect. In `sim_cpg()`, $L = (\text{bmi}_z,
\text{sex})$, where $\text{bmi}_z$ is a standardized body-mass index (a
$z$-score), and
\begin{align*}
g_0(L) &= \operatorname{expit}(-1 + 0.7\, \text{bmi}_z^2 + 0.4\, \text{sex}) \ , \\
\overline{Q}_0(A, L) &= 0.5 - A + 0.3\, \text{bmi}_z + 0.8\, \text{bmi}_z^2 +
  0.2\, \text{sex} \ ,
\end{align*}
so the true ATE is $\psi_0 = -1$. A main-terms GLM in $(A, \text{bmi}_z,
\text{sex})$ omits the confounding through $\text{bmi}_z^2$ and is
misspecified; the learner `SL.glm_correct` (in `practical/R/learners.R`)
adds $\text{bmi}_z^2$ and is correctly specified.

```{r}
#| label: sim-cpg
cpg_sim <- sim_cpg()
a_sim <- cpg_sim$A
l_sim <- cpg_sim[, c("bmi_z", "sex")]
head(cpg_sim, 3)
```

### Exercise 1.2 💻 {#sec-ex12}

::: {.callout-note title="TMLE by hand"}
Complete the steps below to compute the TMLE of the ATE of the synthetic
CpG, with correctly specified GLMs for both $g_0$ and $\overline{Q}_0$.
Check that the average of the estimated EIF at the targeted fit is
(numerically) zero.
:::

```{r}
#| label: ex12-skeleton
#| eval: false
# 1. scale the outcome to [0, 1]
y_range <- range(cpg_sim$Y)
y_s <- scale_to_unit(cpg_sim$Y, y_range)

# 2. initial estimates of g_0 and of Q_0 (on the unit scale)
g_n <- estimate_g(a_sim, l_sim, g_lib = "SL.glm_correct")
Q_n <- estimate_Q(y_s, a_sim, l_sim, q_lib = ___)
QA <- bound_precision(Q_n$QA)
Q1 <- bound_precision(Q_n$Q1)
Q0 <- bound_precision(Q_n$Q0)

# 3. the clever covariate, at the observed A, at A = 1, and at A = 0
H_A <- ___
H_1 <- 1 / g_n
H_0 <- -1 / (1 - g_n)

# 4. fluctuate: logistic regression on H_A, with the initial fit as offset
fluc <- glm(y_s ~ -1 + H_A + offset(qlogis(QA)), family = quasibinomial())
eps <- coef(fluc)

# 5. update the initial fit, and plug in
Q1_star <- plogis(qlogis(Q1) + eps * ___)
Q0_star <- plogis(qlogis(Q0) + eps * ___)
QA_star <- plogis(qlogis(QA) + eps * H_A)
psi_s <- mean(___)

# 6. the EIF at the targeted fit, and the estimate on the original scale
eif <- diff(y_range) * (H_A * (y_s - QA_star) + Q1_star - Q0_star - psi_s)
c(
  psi = diff(y_range) * psi_s,
  se = sd(eif) / sqrt(nrow(cpg_sim)),
  mean_eif = mean(eif)
)
```

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| label: ex12-solution
# 1. scale the outcome to [0, 1]
y_range <- range(cpg_sim$Y)
y_s <- scale_to_unit(cpg_sim$Y, y_range)

# 2. initial estimates of g_0 and of Q_0 (on the unit scale)
g_n <- estimate_g(a_sim, l_sim, g_lib = "SL.glm_correct")
Q_n <- estimate_Q(y_s, a_sim, l_sim, q_lib = "SL.glm_correct")
QA <- bound_precision(Q_n$QA)
Q1 <- bound_precision(Q_n$Q1)
Q0 <- bound_precision(Q_n$Q0)

# 3. the clever covariate, at the observed A, at A = 1, and at A = 0
H_A <- a_sim / g_n - (1 - a_sim) / (1 - g_n)
H_1 <- 1 / g_n
H_0 <- -1 / (1 - g_n)

# 4. fluctuate: logistic regression on H_A, with the initial fit as offset
fluc <- glm(y_s ~ -1 + H_A + offset(qlogis(QA)), family = quasibinomial())
eps <- coef(fluc)

# 5. update the initial fit, and plug in
Q1_star <- plogis(qlogis(Q1) + eps * H_1)
Q0_star <- plogis(qlogis(Q0) + eps * H_0)
QA_star <- plogis(qlogis(QA) + eps * H_A)
psi_s <- mean(Q1_star - Q0_star)

# 6. the EIF at the targeted fit, and the estimate on the original scale
eif <- diff(y_range) * (H_A * (y_s - QA_star) + Q1_star - Q0_star - psi_s)
c(
  psi = diff(y_range) * psi_s,
  se = sd(eif) / sqrt(nrow(cpg_sim)),
  mean_eif = mean(eif)
)

# the initial (untargeted) plug-in estimate, for comparison
diff(y_range) * mean(Q1 - Q0)
```
With correctly specified GLMs, the initial plug-in estimate is already close
to $-1$, and the targeting step barely moves it ($\epsilon_n$ is small). The
mean of the EIF at the targeted fit is zero up to numerical error: that is
what the fluctuation step is for.

As a check, `tmle_ate()`, given the propensity score estimates and a library
for $\overline{Q}_0$, reproduces the computation above:

```{r}
#| label: ex12-check
set.seed(11249)
g_n <- estimate_g(a_sim, l_sim, "SL.glm_correct")
tmle_ate(cpg_sim$Y, a_sim, l_sim, g_n, "SL.glm_correct")
```
:::
:::

### Exercise 1.3 💻 {#sec-ex13}

::: {.callout-note title="Misspecified GLMs, correct GLMs, and Super Learner"}
a. Using `estimate_g()` and `tmle_ate()`, estimate the ATE of the synthetic
   CpG with all four combinations of misspecified (`"SL.glm"`) and correct
   (`"SL.glm_correct"`) GLMs for $g_0$ and $\overline{Q}_0$. Before running
   the code: which combinations do you expect to give a (nearly) unbiased
   estimate? How does the main-terms regression coefficient compare?
b. In practice, we don't know which GLM is correct. Re-estimate the ATE with
   Super Learner for both nuisance parameters, using the library below,
   which includes the correct GLM *among* other candidates: the main-terms
   GLM, lasso and ridge regression (`glmnet`), MARS (`earth`), and a
   generalized additive model (`gam`). Which learners does the Super Learner
   weight most heavily?
:::

```{r}
#| label: ex13-skeleton
#| eval: false
# (a) GLMs
g_wrong <- estimate_g(a_sim, l_sim, "SL.glm")
g_right <- estimate_g(a_sim, l_sim, ___)
rbind(
  both_wrong = tmle_ate(cpg_sim$Y, a_sim, l_sim, g_wrong, "SL.glm"),
  both_right = tmle_ate(cpg_sim$Y, a_sim, l_sim, g_right, "SL.glm_correct"),
  q_wrong_g_right = tmle_ate(___),
  q_right_g_wrong = tmle_ate(___)
)
coef(lm(Y ~ A + bmi_z + sex, data = cpg_sim))[["A"]]

# (b) Super Learner
sl_lib <- c(
  "SL.mean",
  "SL.glm",
  "SL.glm_correct",
  "SL.glmnet",
  "SL.ridge",
  "SL.earth",
  "SL.gam"
)
g_sl <- estimate_g(___)
tmle_ate(cpg_sim$Y, a_sim, l_sim, g_sl, ___)
round(estimate_Q(cpg_sim$Y, a_sim, l_sim, sl_lib)$coef, 2)
```

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| label: ex13-solution
# (a) GLMs
g_wrong <- estimate_g(a_sim, l_sim, "SL.glm")
g_right <- estimate_g(a_sim, l_sim, "SL.glm_correct")
rbind(
  both_wrong = tmle_ate(cpg_sim$Y, a_sim, l_sim, g_wrong, "SL.glm"),
  both_right = tmle_ate(cpg_sim$Y, a_sim, l_sim, g_right, "SL.glm_correct"),
  q_wrong_g_right = tmle_ate(cpg_sim$Y, a_sim, l_sim, g_right, "SL.glm"),
  q_right_g_wrong = tmle_ate(cpg_sim$Y, a_sim, l_sim, g_wrong, "SL.glm_correct")
)
coef(lm(Y ~ A + bmi_z + sex, data = cpg_sim))[["A"]]

# (b) Super Learner
sl_lib <- c(
  "SL.mean",
  "SL.glm",
  "SL.glm_correct",
  "SL.glmnet",
  "SL.ridge",
  "SL.earth",
  "SL.gam"
)
g_sl <- estimate_g(a_sim, l_sim, sl_lib)
tmle_ate(cpg_sim$Y, a_sim, l_sim, g_sl, sl_lib)
round(estimate_Q(cpg_sim$Y, a_sim, l_sim, sl_lib)$coef, 2)
```
a. By double robustness, every combination except "both wrong" is
   consistent. With both GLMs misspecified, the estimate is far from $-1$:
   the omitted $\text{bmi}_z^2$ confounds the comparison, just as it does for
   the main-terms regression coefficient. With only $g_n$ correct, the
   estimate is nearly unbiased but noisier: the targeting step then does all
   the work, through the weights $H_n$, which are large for some people.
b. The Super Learner puts most of its weight on `SL.glm_correct`: by
   cross-validation, it finds the correctly specified GLM among the
   candidates and recovers the truth. Had no candidate been correct, it
   would still pick the best-predicting combination, and double robustness
   only needs one of $g_n$ and $\overline{Q}_n$ to be (close to) right.
:::
:::

## Back to the real data

For the methylation data, we do not know the correct model, so we rely on the
Super Learner. We drop `SL.glm_correct`, which is specific to the synthetic
CpG, and `SL.gam`, whose backfitting fails on the nearly collinear cell-type
proportions (which sum to about one).

### Exercise 1.4 💻 {#sec-ex14}

::: {.callout-note title="The linear model, and *AHRR*"}
a. Fit the working linear model of Exercise 1.1(a) to each CpG, collecting
   $\beta_{n,1}$, its standard error, and its $t$-statistic.
b. Notice that $g_0(L)$ does not depend on $b$: estimate it **once**, with
   the Super Learner.
c. Estimate the ATE of smoking on `cg05575921` (*AHRR*) by TMLE, with the
   Super Learner and with `q_lib = "SL.glm"`. Compare both to the linear
   model's $\beta_{n,1}$. Why is the `"SL.glm"` version so close to it?
:::

```{r}
#| label: ex14-skeleton
#| eval: false
# (a) the linear-model baseline
lm_one <- function(cpg) {
  fit <- lm(reformulate(c("smoker", covars), response = cpg), data = methy)
  coef(summary(fit))[___, c("Estimate", "Std. Error", "t value")]
}
lm_res <- t(sapply(cpgs, lm_one))
head(lm_res[order(-abs(lm_res[, "t value"])), ])

# (b) the propensity score, once
real_lib <- setdiff(sl_lib, c("SL.glm_correct", "SL.gam"))
g_n <- estimate_g(___, ___, real_lib)

# (c) one CpG
rbind(sl = tmle_ate(methy$cg05575921, A, L, g_n, real_lib), glm = tmle_ate(___))
lm_res["cg05575921", ]
```

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| label: ex14-solution
# (a) the linear-model baseline
lm_one <- function(cpg) {
  fit <- lm(reformulate(c("smoker", covars), response = cpg), data = methy)
  coef(summary(fit))["smoker", c("Estimate", "Std. Error", "t value")]
}
lm_res <- t(sapply(cpgs, lm_one))
head(lm_res[order(-abs(lm_res[, "t value"])), ])

# (b) the propensity score, once
real_lib <- setdiff(sl_lib, c("SL.glm_correct", "SL.gam"))
g_n <- estimate_g(A, L, real_lib)
summary(g_n)

# (c) one CpG
rbind(
  sl = tmle_ate(methy$cg05575921, A, L, g_n, real_lib),
  glm = tmle_ate(methy$cg05575921, A, L, g_n, "SL.glm")
)
lm_res["cg05575921", ]
```
The *AHRR* CpG is at the top of the linear-model ranking, as in large
epigenome-wide studies of smoking [@joehanes2016epigenetic]. With $\overline{Q}_{n,b}$ from the main-terms GLM, the initial
plug-in estimate $\frac{1}{n}\sum_i \{\overline{Q}_{n,b}(1, L_i) -
\overline{Q}_{n,b}(0, L_i)\}$ is exactly $\beta_{n,1}$ (Exercise 1.1a); the
targeting step then moves it only as much as the data, through $g_n$,
indicate that the linear model is misspecified. When the linear model fits
well, that step is small.
:::
:::

### Exercise 1.5 💻 {#sec-ex15}

::: {.callout-note title="All CpGs"}
Apply `tmle_ate()`, with the Super Learner, to every CpG. (This takes a
minute or two.) Since we will need the variance of the EIF,
$\sigma^2_{n,b}$, recover it from the standard error.
:::

```{r}
#| label: ex15-skeleton
#| eval: false
tmle_res <- t(sapply(cpgs, function(cpg) tmle_ate(___)))
tmle_res <- cbind(tmle_res, s2 = n * tmle_res[, "se"]^2)
head(tmle_res)
```

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| label: ex15-solution
tmle_res <- t(sapply(cpgs, function(cpg) {
  tmle_ate(methy[[cpg]], A, L, g_n, real_lib)
}))
tmle_res <- cbind(tmle_res, s2 = n * tmle_res[, "se"]^2)
head(tmle_res)
cor(tmle_res[, "psi"], lm_res[, "Estimate"])
```
:::
:::

## Variance moderation

With many biomarkers and few samples, some $\sigma^2_{n,b}$ will be tiny by
chance, producing large test statistics for small effects. The moderated
$t$-statistic of @smyth2004linear borrows strength across biomarkers by
shrinking each variance towards a common value,
$$
\tilde{\sigma}^2_{n,b} = \frac{d_b \, \sigma^2_{n,b} + d_0 \, s_0^2}{d_b +
  d_0}, \qquad
\tilde{t}_{n,b} = \frac{\psi^\star_{n,b}}{\sqrt{\tilde{\sigma}^2_{n,b} / n}},
$$
where $d_b$ are the degrees of freedom of $\sigma^2_{n,b}$, and the prior
degrees of freedom $d_0$ and prior variance $s_0^2$ are estimated from all
$B$ variances by empirical Bayes. $\tilde{t}_{n,b}$ is compared to a
$t$-distribution with $d_b + d_0$ degrees of freedom. `biotmle`'s key
observation [@hejazi2017generalization] is that nothing in this recipe
requires $\sigma^2_{n,b}$ to come from a linear model: it can be the
variance of the EIF.

The function `moderate(s2, df)` (in `practical/R/moderate.R`) implements this
shrinkage, reproducing `limma::squeezeVar()`; it returns `s2_tilde`, `d0`,
and `s0_2`.

### Exercise 1.6 ✏️ {#sec-ex16}

::: {.callout-note title="How much shrinkage?"}
a. What does $\tilde{\sigma}^2_{n,b}$ become as $d_0 \to 0$? As
   $d_0 \to \infty$?
b. Here, $d_b = n - 1 = 252$. Suppose empirical Bayes estimates $d_0 \approx
   5$. What fraction of $\tilde{\sigma}^2_{n,b}$ comes from the prior? When
   would moderation matter more?
c. Why might the EIF-based variance $\sigma^2_{n,b}$ be *more* erratic
   than the linear model's residual variance, making moderation especially
   valuable?
:::

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
a. As $d_0 \to 0$, $\tilde{\sigma}^2_{n,b} \to \sigma^2_{n,b}$: no
   moderation, an ordinary $t$-statistic. As $d_0 \to \infty$,
   $\tilde{\sigma}^2_{n,b} \to s_0^2$: every biomarker shares one variance,
   and tests rank biomarkers by effect size alone.
b. The weight on the prior is $d_0 / (d_b + d_0) = 5 / 257 \approx 2\%$, so
   moderation barely changes anything at $n = 253$. It matters when $n$ is
   small (say, $n = 10$–$30$, as in many expression studies), where
   $d_b$ is comparable to $d_0$.
c. The EIF involves $H_n$, i.e., $1 / g_n(L)$, which can be large for a few
   people (as in Exercise 1.3a). It also involves flexible
   $\overline{Q}_{n,b}$ fits, whose residuals vary across CpGs. The EIF
   variance estimator is only *asymptotically* valid: in small samples, it
   can be too small for some CpGs (false positives) and too large for others.
   Shrinkage stabilizes it without giving up model-agnostic inference.
:::
:::

### Exercise 1.7 💻 {#sec-ex17}

::: {.callout-note title="Moderated tests and multiple testing"}
a. Moderate the EIF variances, compute moderated $t$-statistics and
   $p$-values, and adjust for multiple testing with the
   Benjamini–Hochberg procedure [@benjamini1995controlling].
b. *(Optional.)* Applying the *same* function to the linear models' residual
   variances gives `limma`'s moderated $t$-statistic. Do so, and compare the
   two lists of discoveries.
:::

```{r}
#| label: ex17-skeleton
#| eval: false
# (a) biotmle: moderate the EIF variances
mod_eif <- moderate(tmle_res[, "s2"], df = ___)
t_eif <- tmle_res[, "psi"] / sqrt(___ / n)
p_eif <- 2 * pt(-abs(t_eif), df = n - 1 + mod_eif$d0)
q_eif <- p.adjust(p_eif, method = "BH")

# (b) limma: moderate the residual variances of the linear models
df_lm <- n - length(covars) - 2
...
```

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| label: ex17-solution
# (a) biotmle: moderate the EIF variances
mod_eif <- moderate(tmle_res[, "s2"], df = n - 1)
t_eif <- tmle_res[, "psi"] / sqrt(mod_eif$s2_tilde / n)
p_eif <- 2 * pt(-abs(t_eif), df = n - 1 + mod_eif$d0)
q_eif <- p.adjust(p_eif, method = "BH")
c(d0 = mod_eif$d0, s0_2 = mod_eif$s0_2)

# (b) limma: moderate the residual variances of the linear models
df_lm <- n - length(covars) - 2
sigma_lm <- sapply(cpgs, function(cpg) {
  sigma(lm(reformulate(c("smoker", covars), response = cpg), data = methy))
})
unscaled_se <- lm_res[, "Std. Error"] / sigma_lm
mod_lm <- moderate(sigma_lm^2, df = df_lm)
t_lm <- lm_res[, "Estimate"] / (unscaled_se * sqrt(mod_lm$s2_tilde))
p_lm <- 2 * pt(-abs(t_lm), df = df_lm + mod_lm$d0)
q_lm <- p.adjust(p_lm, method = "BH")

table(biotmle = q_eif < 0.05, limma = q_lm < 0.05)
```

```{r}
#| label: ex17-volcano
#| fig-cap: "Volcano plot of the variance-moderated TMLE estimates."
res <- data.frame(cpg = cpgs, psi = tmle_res[, "psi"], q = q_eif)
ggplot(res, aes(psi, -log10(q), colour = q < 0.05)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_text(
    data = subset(res, cpg == "cg05575921"),
    aes(label = "AHRR"),
    nudge_x = 0.15,
    colour = "black"
  ) +
  labs(
    x = "Estimated ATE of smoking (M-value scale)",
    y = expression(-log[10](q)),
    colour = "FDR < 5%"
  ) +
  theme_bw()
```

The two procedures largely agree here: the linear model seems a reasonable
working model for most CpGs, and $n$ is large enough that moderation is
mild. Where they disagree is precisely where the linear model's assumptions
are doing work that the data cannot verify.
:::
:::

### Exercise 1.8 ✏️ *(stretch)* {#sec-ex18}

::: {.callout-note title="Two-stage analyses"}
These 300 CpGs were themselves selected: half passed a screen applied to
all 472,678 CpGs measured. The published analysis
[@hejazi2017generalization] screens with `limma`, runs `biotmle` on the 2,537
CpGs that pass, and adjusts $p$-values as if all 472,678 had been tested
(setting $p = 1$ for those screened out) [@tuglus2009modified]. Why is
adjusting only over the CpGs that pass the screen not enough?
:::

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
The screen used the same data, and kept CpGs *because* their association
with smoking looked strong. Testing only the survivors, and correcting only
for their number, ignores the selection: the error rate would be computed
over a set of hypotheses chosen to look significant. Counting every CpG
measured in the multiplicity correction accounts for the first stage.
:::
:::
