# Part 1: Biomarker discovery as causal effect estimation {#sec-biotmle}

```{r}
#| label: setup-biotmle
#| include: false
library(SuperLearner)
library(ggplot2)
source("practical/R/moderate.R")
methy <- read.csv("practical/data/su2016_methylation.csv.gz")
set.seed(11249)
```

## The data: smoking and DNA methylation

@su2016distinct measured DNA methylation in whole blood, at roughly 470,000
CpG sites, in 253 people: 172 who had ever smoked and 81 who never had. We
use their data to ask: **which CpG sites are affected by smoking?**

The data structure for one person is $O = (L, A, Y)$, where

- $L$ are baseline covariates: sex, age, race, and the estimated proportions
  of six leukocyte types (`CD8T`, `CD4T`, `NK`, `Bcell`, `Mono`, `Gran`) —
  blood cell composition differs by smoking status and strongly affects
  methylation, so it is a confounder;
- $A \in \{0, 1\}$ is smoking status (`smoker`: 1 = ever, 0 = never);
- $Y = (Y_b : b = 1, \ldots, B)$ are the methylation levels (M-values) at
  each CpG site $b$.

To keep things fast, the file contains $B = 300$ CpG sites: 150 chosen at
random among those flagged by a `limma` screen in the full data (including
four well-known smoking-related CpGs, e.g., `cg05575921` in the *AHRR* gene),
and 150 chosen at random among those that were not.

```{r}
#| label: look-data
covars <- c("sex", "age", "race", "CD8T", "CD4T", "NK", "Bcell", "Mono", "Gran")
cpgs <- grep("^cg", names(methy), value = TRUE)
A <- methy$smoker
L <- methy[, covars]
n <- nrow(methy)

dim(methy)
table(smoker = A)
head(methy[, c("smoker", covars, cpgs[1:2])], 3)
```

## The target parameter

For each CpG $b$, our target is the ATE of smoking on methylation, which
(under no unmeasured confounding and positivity) is identified by
$$
\psi_b = \E_L\big[ Q_b(1, L) - Q_b(0, L) \big], \qquad
Q_b(a, l) \coloneqq \E[Y_b \mid A = a, L = l].
$$
This is a *variable importance measure*: it ranks CpG sites by how much
smoking would change their methylation, without committing to a model for
$Q_b$.

### Exercise 1.1 ✏️ — When is a regression coefficient the ATE? {#sec-ex11}

The usual approach (as in `limma`) fits, for each CpG, the working linear
model $Q_b(a, l) = \beta_0 + \beta_1 a + \gamma^\top l$, and tests
$\beta_1 = 0$.

a. Suppose this linear model is correct. Show that $\psi_b = \beta_1$.
b. Now suppose the effect of smoking varies with age, so that the correct
   model is $Q_b(a, l) = \beta_0 + \beta_1 a + \beta_2 a \cdot
   \text{age} + \gamma^\top l$. What is $\psi_b$ now? Is $\beta_1$ still
   the ATE?
c. In (b), what would $\beta_1$ in the *main-terms* model of (a) estimate?
   Is there a study design under which it would still estimate the ATE?

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
a. $Q_b(1, l) - Q_b(0, l) = (\beta_0 + \beta_1 + \gamma^\top l) - (\beta_0 +
   \gamma^\top l) = \beta_1$ for every $l$, so averaging over $L$ gives
   $\psi_b = \beta_1$.
b. Now $Q_b(1, l) - Q_b(0, l) = \beta_1 + \beta_2 \, \text{age}$, so
   $\psi_b = \beta_1 + \beta_2 \E[\text{age}]$. The coefficient $\beta_1$ is
   the effect at age 0 — not a quantity anyone cares about.
c. The main-terms coefficient converges to (roughly) a weighted average of
   the age-specific effects, with weights proportional to the conditional
   variance of smoking, $\V(A \mid L)$ — people whose smoking status is hard
   to predict from $L$ count more. It is not $\psi_b$ in general. If $A$ were randomized ($A \indep
   L$), the main-terms coefficient would be consistent for the ATE even
   though the model is wrong. Smoking is not randomized, so here we need an
   estimator whose validity does not hinge on the form of $Q_b$.

**Take-away:** $\psi_b$ is defined without reference to any model; $\beta_1$
*is* $\psi_b$ only if the linear model is right (or $A$ is randomized).
:::
:::

### Exercise 1.2 💻 — The linear-model baseline {#sec-ex12}

Fit the working linear model of Exercise 1.1(a) to each CpG and collect the
estimated $\beta_1$, its standard error, and its $t$-statistic.

```{r}
#| label: ex12-skeleton
#| eval: false
lm_one <- function(cpg) {
  fit <- lm(reformulate(c("smoker", covars), response = cpg), data = methy)
  coef(summary(fit))[___, c("Estimate", "Std. Error", "t value")]
}
lm_res <- t(sapply(cpgs, lm_one))
head(lm_res[order(-abs(lm_res[, "t value"])), ])
```

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| label: ex12-solution
lm_one <- function(cpg) {
  fit <- lm(reformulate(c("smoker", covars), response = cpg), data = methy)
  coef(summary(fit))["smoker", c("Estimate", "Std. Error", "t value")]
}
lm_res <- t(sapply(cpgs, lm_one))
head(lm_res[order(-abs(lm_res[, "t value"])), ])
```
The *AHRR* CpG `cg05575921` is at the top: smokers have markedly lower
methylation there, a finding replicated in dozens of studies.
:::
:::

## An efficient, model-agnostic estimator

From the lecture, the efficient influence function (EIF) of $\psi_b$ is
$$
D_b(O) = \left( \frac{A}{g(L)} - \frac{1 - A}{1 - g(L)} \right)
  \big(Y_b - Q_b(A, L)\big) + Q_b(1, L) - Q_b(0, L) - \psi_b,
$$
where $g(L) = \Pr(A = 1 \mid L)$ is the propensity score. Given estimates
$\hat{g}$ and $\hat{Q}_b$ — from any machine learning method, e.g., the
Super Learner [@vdl2007super] — the *one-step* (or augmented IPW) estimator
is the average of the uncentered EIF,
$$
\hat{\psi}_b = \frac{1}{n} \sum_{i = 1}^n \left\{
  \left( \frac{A_i}{\hat{g}(L_i)} - \frac{1 - A_i}{1 - \hat{g}(L_i)} \right)
  \big(Y_{b,i} - \hat{Q}_b(A_i, L_i)\big)
  + \hat{Q}_b(1, L_i) - \hat{Q}_b(0, L_i) \right\},
$$
with standard error $\sqrt{\hat{\sigma}^2_b / n}$, where $\hat{\sigma}^2_b$
is the empirical variance of the estimated EIF. (`biotmle` uses TMLE rather
than the one-step estimator; the two are asymptotically equivalent, and the
one-step estimator takes fewer lines of code.)

### Exercise 1.3 💻 — The propensity score, once {#sec-ex13}

Notice that $g(L)$ does not depend on $b$. Estimate it **once**, with a
Super Learner, and bound the estimates away from 0 and 1.

```{r}
#| label: ex13-skeleton
#| eval: false
sl_lib <- c("SL.mean", "SL.glm", "SL.glmnet", "SL.earth")
g_fit <- SuperLearner(Y = ___, X = ___, family = binomial(),
                      SL.library = sl_lib, cvControl = list(V = 5))
g_hat <- pmin(pmax(g_fit$SL.predict[, 1], 0.01), 0.99)
summary(g_hat)
```

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| label: ex13-solution
sl_lib <- c("SL.mean", "SL.glm", "SL.glmnet", "SL.earth")
g_fit <- SuperLearner(Y = A, X = L, family = binomial(),
                      SL.library = sl_lib, cvControl = list(V = 5))
g_hat <- pmin(pmax(g_fit$SL.predict[, 1], 0.01), 0.99)
summary(g_hat)
g_fit$coef
```
The propensity scores lie well inside $(0, 1)$: no practical positivity
problems.
:::
:::

### Exercise 1.4 💻 — One CpG, by hand {#sec-ex14}

Complete the function below, which estimates $Q_b$ with a Super Learner,
predicts it at $(A_i, L_i)$, $(1, L_i)$, and $(0, L_i)$, and returns the
one-step estimate and the variance of the EIF. Apply it to `cg05575921`.

```{r}
#| label: ex14-skeleton
#| eval: false
aipw <- function(y, q_lib = sl_lib) {
  X <- data.frame(A = A, L)
  X1 <- transform(X, A = 1)
  X0 <- transform(X, A = 0)
  q_fit <- SuperLearner(Y = y, X = X, newX = rbind(X, X1, X0),
                        SL.library = q_lib, cvControl = list(V = 5))
  q_hat <- matrix(q_fit$SL.predict, ncol = 3)  # columns: Q(A,L), Q(1,L), Q(0,L)
  eif <- ___
  c(psi = mean(eif), s2 = var(eif))
}
ahrr <- aipw(methy$cg05575921)
c(psi = ahrr[["psi"]], se = sqrt(ahrr[["s2"]] / n))
lm_res["cg05575921", ]
```

Then answer:

a. How does $\hat{\psi}_b$ compare to the linear model's $\hat{\beta}_1$?
b. Re-run with `q_lib = "SL.glm"`, i.e., using the linear model for
   $Q_b$. What do you get, and why?

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| label: ex14-solution
aipw <- function(y, q_lib = sl_lib) {
  X <- data.frame(A = A, L)
  X1 <- transform(X, A = 1)
  X0 <- transform(X, A = 0)
  q_fit <- SuperLearner(Y = y, X = X, newX = rbind(X, X1, X0),
                        SL.library = q_lib, cvControl = list(V = 5))
  q_hat <- matrix(q_fit$SL.predict, ncol = 3)  # columns: Q(A,L), Q(1,L), Q(0,L)
  eif <- (A / g_hat - (1 - A) / (1 - g_hat)) * (y - q_hat[, 1]) +
    q_hat[, 2] - q_hat[, 3]
  c(psi = mean(eif), s2 = var(eif))
}
ahrr <- aipw(methy$cg05575921)
c(psi = ahrr[["psi"]], se = sqrt(ahrr[["s2"]] / n))
lm_res["cg05575921", ]

ahrr_glm <- aipw(methy$cg05575921, q_lib = "SL.glm")
c(psi = ahrr_glm[["psi"]], se = sqrt(ahrr_glm[["s2"]] / n))
```
a. Very close: for this CpG, the linear model appears to be a decent
   approximation of $Q_b$.
b. With $\hat{Q}_b$ from the main-terms linear model, the plug-in part
   $\frac{1}{n}\sum_i \{\hat{Q}_b(1, L_i) - \hat{Q}_b(0, L_i)\}$ is exactly
   $\hat{\beta}_1$ (Exercise 1.1a). The one-step estimator adds a correction,
   $\frac{1}{n}\sum_i (\ldots)(Y_{b,i} - \hat{Q}_b(A_i, L_i))$, which is
   small when the linear model fits well — and which *repairs* the estimate
   when it does not. Either way, the estimator stays valid as long as
   $\hat{g}$ or $\hat{Q}_b$ is consistent (double robustness).
:::
:::

### Exercise 1.5 💻 — All CpGs {#sec-ex15}

Apply `aipw()` to every CpG. (This takes about a minute.)

```{r}
#| label: ex15-skeleton
#| eval: false
eif_res <- t(sapply(cpgs, function(cpg) aipw(___)))
head(eif_res)
```

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| label: ex15-solution
eif_res <- t(sapply(cpgs, function(cpg) aipw(methy[[cpg]])))
head(eif_res)
cor(eif_res[, "psi"], lm_res[, "Estimate"])
```
:::
:::

## Variance moderation

With many biomarkers and few samples, some $\hat{\sigma}^2_b$ will be tiny by
chance, producing large test statistics for small effects. The moderated
$t$-statistic of @smyth2004linear borrows strength across biomarkers by
shrinking each variance towards a common value,
$$
\tilde{\sigma}^2_b = \frac{d_b \, \hat{\sigma}^2_b + d_0 \, s_0^2}{d_b + d_0},
\qquad
\tilde{t}_b = \frac{\hat{\psi}_b}{\sqrt{\tilde{\sigma}^2_b / n}},
$$
where $d_b$ are the degrees of freedom of $\hat{\sigma}^2_b$, and the prior
degrees of freedom $d_0$ and prior variance $s_0^2$ are estimated from all
$B$ variances by empirical Bayes. $\tilde{t}_b$ is compared to a
$t$-distribution with $d_b + d_0$ degrees of freedom. `biotmle`'s key
observation [@hejazi2017generalization] is that nothing in this recipe
requires $\hat{\sigma}^2_b$ to come from a linear model: it can be the
variance of the EIF.

The function `moderate(s2, df)` (in `practical/R/moderate.R`) implements this
shrinkage, reproducing `limma::squeezeVar()`; it returns `s2_tilde`, `d0`,
and `s0_2`.

### Exercise 1.6 ✏️ — How much shrinkage? {#sec-ex16}

a. What does $\tilde{\sigma}^2_b$ become as $d_0 \to 0$? As
   $d_0 \to \infty$?
b. Here, $d_b = n - 1 = 252$. Suppose empirical Bayes estimates $d_0 \approx
   5$. What fraction of $\tilde{\sigma}^2_b$ comes from the prior? When
   would moderation matter more?
c. Why might the EIF-based variance $\hat{\sigma}^2_b$ be *more* erratic
   than the linear model's residual variance, making moderation especially
   valuable for `biotmle`?

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
a. As $d_0 \to 0$, $\tilde{\sigma}^2_b \to \hat{\sigma}^2_b$: no
   moderation, an ordinary $t$-statistic. As $d_0 \to \infty$,
   $\tilde{\sigma}^2_b \to s_0^2$: every biomarker shares one variance, and
   tests rank biomarkers by effect size alone.
b. The weight on the prior is $d_0 / (d_b + d_0) = 5 / 257 \approx 2\%$, so
   moderation barely changes anything at $n = 253$. It matters when $n$ is
   small (say, $n = 10$–$30$, as in many expression studies), where
   $d_b$ is comparable to $d_0$.
c. The EIF involves $1 / \hat{g}(L)$, which can be large for a few people,
   and flexible $\hat{Q}_b$ fits whose residuals vary across CpGs; the EIF
   variance estimator is only *asymptotically* valid, and in small samples
   it can be too small for some CpGs (false positives) and too large for
   others. Shrinkage stabilizes it without giving up model-agnostic
   inference.
:::
:::

### Exercise 1.7 💻 — Moderated tests and multiple testing {#sec-ex17}

a. Moderate the EIF variances, compute moderated $t$-statistics and
   $p$-values, and adjust for multiple testing with the
   Benjamini–Hochberg procedure [@benjamini1995controlling].
b. Applying the *same* function to the linear models' residual variances
   gives `limma`'s moderated $t$-statistic. Do so, and compare the two lists
   of discoveries.

```{r}
#| label: ex17-skeleton
#| eval: false
# (a) biotmle: moderate the EIF variances
mod_eif <- moderate(eif_res[, "s2"], df = ___)
t_eif <- eif_res[, "psi"] / sqrt(___ / n)
p_eif <- 2 * pt(-abs(t_eif), df = n - 1 + mod_eif$d0)
q_eif <- p.adjust(p_eif, method = "BH")

# (b) limma: moderate the residual variances of the linear models
df_lm <- n - length(covars) - 2
s2_lm <- (lm_res[, "Std. Error"] / ___)^2   # hint: SE = sigma * (unscaled SE)
...
```

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| label: ex17-solution
# (a) biotmle: moderate the EIF variances
mod_eif <- moderate(eif_res[, "s2"], df = n - 1)
t_eif <- eif_res[, "psi"] / sqrt(mod_eif$s2_tilde / n)
p_eif <- 2 * pt(-abs(t_eif), df = n - 1 + mod_eif$d0)
q_eif <- p.adjust(p_eif, method = "BH")
c(d0 = mod_eif$d0, s0_2 = mod_eif$s0_2)

# (b) limma: moderate the residual variances of the linear models
df_lm <- n - length(covars) - 2
sigma_lm <- sapply(cpgs, function(cpg) {
  sigma(lm(reformulate(c("smoker", covars), response = cpg), data = methy))
})
unscaled_se <- lm_res[, "Std. Error"] / sigma_lm
mod_lm <- moderate(sigma_lm^2, df = df_lm)
t_lm <- lm_res[, "Estimate"] / (unscaled_se * sqrt(mod_lm$s2_tilde))
p_lm <- 2 * pt(-abs(t_lm), df = df_lm + mod_lm$d0)
q_lm <- p.adjust(p_lm, method = "BH")

table(biotmle = q_eif < 0.05, limma = q_lm < 0.05)
```

```{r}
#| label: ex17-volcano
#| fig-cap: "Volcano plot of the variance-moderated, EIF-based ATE estimates."
res <- data.frame(cpg = cpgs, psi = eif_res[, "psi"], q = q_eif)
ggplot(res, aes(psi, -log10(q), colour = q < 0.05)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_text(data = subset(res, cpg == "cg05575921"), aes(label = "AHRR"),
            nudge_x = 0.15, colour = "black") +
  labs(x = "Estimated ATE of smoking (M-value scale)",
       y = expression(-log[10](q)), colour = "FDR < 5%") +
  theme_bw()
```

The two procedures largely agree here: the linear model seems a reasonable
working model for most CpGs, and $n$ is large enough that moderation is
mild. Where they disagree is precisely where the linear model's assumptions
are doing work that the data cannot verify.
:::
:::

### Exercise 1.8 ✏️ — *Stretch:* two-stage analyses

These 300 CpGs did not come out of nowhere: half passed a screen applied to
all 472,678 CpGs measured. The published analysis
[@hejazi2017generalization] screens with `limma`, runs `biotmle` on the 2,537
CpGs that pass, and adjusts $p$-values as if all 472,678 had been tested
(setting $p = 1$ for those screened out) [@tuglus2009modified]. Why is
adjusting only over the CpGs that pass the screen not enough?

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
The screen used the same data, and kept CpGs *because* their association
with smoking looked strong. Testing only the survivors, and correcting only
for their number, ignores the selection: the error rate would be computed
over a set of hypotheses chosen to look significant. Counting every CpG
measured in the multiplicity correction accounts for the first stage.
:::
:::
