# Part 2: Stochastic-interventional vaccine efficacy {#sec-sve}

```{r}
#| label: setup-sve
#| include: false
library(lmtp)
library(ggplot2)
source("practical/R/sim_vaccine_trial.R")
set.seed(72104)
```

## The question

In vaccine trials, an *immune correlate of protection* is a marker of the
vaccine-induced immune response $S$ (e.g., neutralizing antibody titer)
that tracks protection against the outcome $Y$ (e.g., infection)
[@plotkin2012nomenclature]. Correlates help us understand *how* a vaccine
protects and can serve as surrogate endpoints for future trials
[@prentice1989surrogate; @gilbert2024four]. The question from the lecture:

> How would the risk of infection have changed had the vaccine induced
> higher, or lower, antibody titers?

The vaccine $A$ is randomized, but $S$ is not: it is a *post-treatment*
variable on the pathway $A \to S \to Y$, a mediator. Questions about
intervening on $S$ are questions about mediation.

## The data: a simulated trial with a case-cohort design

We simulate a trial of 10,000 people (in the real trials, the data are not
public). The complete data on one person are $X = (L, A, S, Y)$:

- $L$: standardized `age`, a `high_risk` behaviour indicator, `sex`;
- $A$: vaccine (1) or placebo (0), randomized 1:1;
- $S$: $\log_{10}$ neutralizing antibody titer ($S = 0$, the assay floor,
  in the placebo arm);
- $Y$: infection by the end of follow-up.

Assaying blood samples is expensive, so $S$ is measured only in a
*phase-two* sample ($B = 1$): every infected person (case), plus a random
subcohort of non-cases, sampled with probability 0.25 if `high_risk` and
0.15 otherwise — a case-cohort design [@prentice1986case], a special case
of two-phase sampling [@white1982two; @breslow2003large]. The sampling
probabilities $\pi_B(Y, L) = \Pr(B = 1 \mid Y, L)$ are known by design and
recorded as `pi_B`.

```{r}
#| label: sim-data
trial <- sim_vaccine_trial()
head(trial)
with(trial, table(vaccine = A, infected = Y))
with(trial, table(vaccine = A, phase_two = B))
```

We focus on vaccine recipients in the phase-two sample:

```{r}
#| label: vax-data
vax <- subset(trial, A == 1 & B == 1)
nrow(vax)
mean(vax$Y)   # cases are heavily over-represented!
```

## The target parameter

A *shift intervention* replaces each vaccinee's titer $S$ with $d(S, L) = S
+ \delta$ [@diaz2012population; @haneuse2013estimation]. The counterfactual
risk of infection under this shift, among vaccinees, is identified by
$$
\psi_\delta = \E_L \left[ \int Q(s + \delta, L) \, g_S(s \mid L) \, ds
\right], \qquad Q(s, l) \coloneqq \Pr(Y = 1 \mid S = s, A = 1, L = l),
$$
where $g_S(s \mid L)$ is the conditional density of $S$ among vaccinees,
under no unmeasured confounding of $S$ and $Y$ and positivity. The
**stochastic-interventional vaccine efficacy** [@hejazi2020efficient;
@gilbert2021covpn]
compares this risk to the risk in the placebo arm:
$$
\text{SVE}(\delta) = 1 - \frac{\psi_\delta}{\Pr(Y = 1 \mid A = 0)}.
$$
At $\delta = 0$, this is the usual vaccine efficacy; as $\delta$ varies, it
traces out how efficacy would change with the vaccine's immunogenicity. It was
one of the immune correlates analyses of the COVE trial of the Moderna
mRNA-1273 COVID-19 vaccine [@gilbert2021covpn-cove; @hejazi2023stochastic;
@huang2023stochastic]. Unlike the controlled vaccine efficacy from the keynote
[@gilbert2022controlled], which sets every vaccinee's titer to the *same*
value $s$, a shift moves each person's titer from where it naturally fell.

### Exercise 2.1 ✏️ — Shifts, linear models, and sampling {#sec-ex21}

a. Suppose the outcome model is linear: $Q(s, l) = \beta_0 + \beta_1 s +
   \gamma^\top l$. Show that $\psi_\delta - \psi_0 = \beta_1 \delta$. What
   does $\text{SVE}(\delta)$ look like as a function of $\delta$?
b. What goes wrong with the linear model as $\delta$ grows? (Hint: $\psi$ is
   a probability.)
c. Among the phase-two vaccinees, what is the share of cases, compared to
   the full vaccine arm? What would happen if we estimated $\psi_\delta$
   from `vax` as if it were a random sample of vaccinees? How do the
   weights $1 / \pi_B(Y, L)$ fix this?
d. Positivity for a shift intervention requires that, for each $l$, the
   shifted value $s + \delta$ is a value of $S$ that could occur among
   people with $L = l$. Why is this harder to satisfy for large $|\delta|$?

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
a. $\psi_\delta = \E[\beta_0 + \beta_1 (S + \delta) + \gamma^\top L] =
   \psi_0 + \beta_1 \delta$. So $\text{SVE}(\delta) = 1 - (\psi_0 + \beta_1
   \delta) / p_0$ is a straight line in $\delta$, with slope $-\beta_1 /
   p_0$: the linear model's $\beta_1$ *is* the effect of a unit shift, just
   as $\beta_1$ was the ATE in Part 1.
b. A straight line eventually leaves $[0, 1]$: for large positive $\delta$
   (with $\beta_1 < 0$), the predicted risk becomes negative and the SVE
   exceeds 100%; the model also forces each additional unit of titer to
   buy the same reduction in risk, whatever the starting titer.
c. All cases are sampled but only 15–25% of non-cases, so cases are hugely
   over-represented (compare `mean(vax$Y)` with the infection rate in the
   vaccine arm). Ignoring this would badly overestimate risk. Weighting each
   phase-two person by $1 / \pi_B(Y, L)$ makes them stand in for the
   $1 / \pi_B$ people like them in the full trial (inverse probability
   weighting), recovering the full-cohort target.
d. Shifting by $\delta$ asks about titers $s + \delta$; for large
   $|\delta|$, many shifted titers fall outside the range observed among
   similar people (e.g., below the assay floor, or far above any observed
   response), so the data contain no information about risk there and any
   estimate is an extrapolation.
:::
:::

## Estimation with `lmtp`

The `lmtp` package [@diaz2023nonparametric] implements estimators of the
effects of *modified treatment policies*, of which our shift is one.
`lmtp_tmle()` estimates $\psi_\delta$ by targeted maximum likelihood, fitting
$Q$ and the needed density ratio with a Super Learner. It accepts sampling
weights through `weights`, which gives the inverse-probability-weighted
version of the estimator for two-phase designs [@rose2011targeted].^[The
estimator in the `txshift` package [@hejazi2020txshift] also *augments* the weighted estimator with a model for the sampling mechanism, using information on the whole cohort
to gain efficiency [@breslow2009using; @hejazi2020efficient]. The
reweighted version is valid here because $\pi_B$ is known by design.]

### Exercise 2.2 💻 — One shift {#sec-ex22}

Estimate $\psi_{0.5}$, the risk among vaccinees had every vaccinee's titer
been $0.5$ $\log_{10}$ units (about 3-fold) higher.

```{r}
#| label: ex22-skeleton
#| eval: false
sl_lib <- c("SL.mean", "SL.glm", "SL.earth")
shift_by <- function(delta) function(data, trt) ___
fit_05 <- lmtp_tmle(
  data = vax, trt = "S", outcome = "Y",
  baseline = c("age", "high_risk", "sex"),
  shift = shift_by(0.5), outcome_type = "binomial",
  learners_outcome = sl_lib, learners_trt = sl_lib, folds = 5,
  weights = ___
)
c(estimate = fit_05$estimate@x, std_error = fit_05$estimate@std_error)
```

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| label: ex22-solution
sl_lib <- c("SL.mean", "SL.glm", "SL.earth")
shift_by <- function(delta) function(data, trt) data[[trt]] + delta
fit_05 <- lmtp_tmle(
  data = vax, trt = "S", outcome = "Y",
  baseline = c("age", "high_risk", "sex"),
  shift = shift_by(0.5), outcome_type = "binomial",
  learners_outcome = sl_lib, learners_trt = sl_lib, folds = 5,
  weights = 1 / vax$pi_B
)
c(estimate = fit_05$estimate@x, std_error = fit_05$estimate@std_error)
```
:::
:::

### Exercise 2.3 💻 — The SVE curve {#sec-ex23}

Estimate $\psi_\delta$ over a grid of shifts, compute $\text{SVE}(\delta)$
using the placebo-arm risk (estimable from the full trial, since $S$ is not
needed), and compare to the truth, which we know because we simulated the
data (`true_sve()`).

```{r}
#| label: ex23-skeleton
#| eval: false
deltas <- c(-0.5, -0.25, 0, 0.25, 0.5, 0.75, 1)
risk_placebo <- mean(trial$Y[trial$A == ___])
psi_hat <- sapply(deltas, function(delta) {
  fit <- lmtp_tmle(___)   # as in Exercise 2.2, with shift_by(delta)
  fit$estimate@x
})
sve <- data.frame(delta = deltas, lmtp = 1 - psi_hat / ___,
                  truth = true_sve(deltas)$sve)
sve
```

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| label: ex23-solution
deltas <- c(-0.5, -0.25, 0, 0.25, 0.5, 0.75, 1)
risk_placebo <- mean(trial$Y[trial$A == 0])
fit_shift <- function(delta, weights = 1 / vax$pi_B) {
  lmtp_tmle(
    data = vax, trt = "S", outcome = "Y",
    baseline = c("age", "high_risk", "sex"),
    shift = shift_by(delta), outcome_type = "binomial",
    learners_outcome = sl_lib, learners_trt = sl_lib, folds = 5,
    weights = weights
  )
}
psi_hat <- sapply(deltas, function(delta) fit_shift(delta)$estimate@x)
sve <- data.frame(delta = deltas, lmtp = 1 - psi_hat / risk_placebo,
                  truth = true_sve(deltas)$sve)
sve
```
:::
:::

### Exercise 2.4 💻 — The linear-model analogue {#sec-ex24}

Now estimate the same curve by *g-computation* with parametric outcome
models, fit by weighted regression to `vax`: (i) the linear probability
model of Exercise 2.1(a), and (ii) a logistic regression. Predict each
vaccinee's risk at $S + \delta$, average (with the weights), and convert to
SVE. Plot all four curves.

```{r}
#| label: ex24-skeleton
#| eval: false
w <- 1 / vax$pi_B
fit_lin <- lm(Y ~ S + age + high_risk + sex, data = vax, weights = w)
fit_logit <- glm(Y ~ S + age + high_risk + sex, data = vax, weights = w,
                 family = quasibinomial())
gcomp_sve <- function(fit, delta) {
  risk <- predict(fit, newdata = transform(vax, S = ___), type = "response")
  1 - weighted.mean(risk, w) / risk_placebo
}
sve$linear <- sapply(deltas, gcomp_sve, fit = fit_lin)
sve$logistic <- sapply(deltas, gcomp_sve, fit = fit_logit)
```

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| label: ex24-solution
w <- 1 / vax$pi_B
fit_lin <- lm(Y ~ S + age + high_risk + sex, data = vax, weights = w)
fit_logit <- glm(Y ~ S + age + high_risk + sex, data = vax, weights = w,
                 family = quasibinomial())
gcomp_sve <- function(fit, delta) {
  risk <- predict(fit, newdata = transform(vax, S = S + delta),
                  type = "response")
  1 - weighted.mean(risk, w) / risk_placebo
}
sve$linear <- sapply(deltas, gcomp_sve, fit = fit_lin)
sve$logistic <- sapply(deltas, gcomp_sve, fit = fit_logit)
round(sve, 3)

# the linear model's slope: SVE(delta) = SVE(0) - beta_1 * delta / p_0
-coef(fit_lin)[["S"]] / risk_placebo
```

```{r}
#| label: ex24-plot
#| fig-cap: "SVE as a function of the shift in log10 titer."
sve_long <- reshape(sve, direction = "long", varying = names(sve)[-1],
                    v.names = "sve", timevar = "method",
                    times = names(sve)[-1], idvar = "delta")
ggplot(sve_long, aes(delta, sve, colour = method)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 1, linetype = "dotted") +
  labs(x = expression(paste("Shift ", delta, " in ", log[10], " titer")),
       y = "SVE", colour = NULL) +
  theme_bw()
```

The linear model forces a straight line and, by $\delta = 1$, claims an
efficacy above 100%: its "risk" is negative. Logistic regression happens to
do well here, because the true model is not far from logistic in the range
of the data — but we only know that because we simulated the data.
`lmtp` makes no such bet.
:::
:::

### Exercise 2.5 💻 — Forgetting the design {#sec-ex25}

Re-estimate $\text{SVE}(0.5)$ with `lmtp_tmle()` **without** the sampling
weights. What happens, and why?

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| label: ex25-solution
fit_unwt <- fit_shift(0.5, weights = NULL)
c(weighted = 1 - fit_05$estimate@x / risk_placebo,
  unweighted = 1 - fit_unwt$estimate@x / risk_placebo,
  truth = true_sve(0.5)$sve)
```
Without weights, the phase-two sample is treated as representative of all
vaccinees, but cases are over-represented about five-fold (15% of the
phase-two vaccinees versus 3% of all vaccinees), so the estimated risk is far
too high and the SVE far too low. The two-phase design must be built into
the estimator.
:::
:::

### Exercise 2.6 💻 — *Stretch:* positivity and realistic shifts

a. Estimate $\text{SVE}(-1)$. Compare to the truth (`true_sve(-1)`). What
   might explain the result? (Hint: what is the range of `vax$S`, and what
   does $S - 1$ mean for people near the bottom of it?)
b. A shift that respects positivity only shifts people whose shifted titer
   remains plausible [@hejazi2020efficient]: $d(s, l) = s + \delta$ if $s
   + \delta \leq u$, and $d(s, l) = s$ otherwise. Implement it with $u =
   \max(S)$ in the phase-two sample and re-estimate the SVE curve. How
   would you interpret this effect?
c. `fit$estimate@std_error` gives the standard error of $\hat{\psi}_\delta$.
   Use the delta method to construct a 95% confidence interval for
   $\text{SVE}(0.5)$, treating the placebo risk as known.

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| label: ex26-solution
# (a) a large negative shift
range(vax$S)
fit_m1 <- fit_shift(-1)
c(estimate = 1 - fit_m1$estimate@x / risk_placebo, truth = true_sve(-1)$sve)

# (b) a positivity-respecting shift, capped at the largest observed titer
u <- max(vax$S)
shift_capped <- function(delta) {
  function(data, trt) ifelse(data[[trt]] + delta <= u, data[[trt]] + delta,
                             data[[trt]])
}
psi_capped <- sapply(deltas, function(delta) {
  lmtp_tmle(
    data = vax, trt = "S", outcome = "Y",
    baseline = c("age", "high_risk", "sex"),
    shift = shift_capped(delta), outcome_type = "binomial",
    learners_outcome = sl_lib, learners_trt = sl_lib, folds = 5,
    weights = 1 / vax$pi_B
  )$estimate@x
})
round(data.frame(delta = deltas, capped = 1 - psi_capped / risk_placebo), 3)

# (c) delta-method CI: SVE = 1 - psi / p0, so SE(SVE) = SE(psi) / p0
est <- fit_05$estimate
(1 - est@x / risk_placebo) +
  c(-1, 0, 1) * qnorm(0.975) * est@std_error / risk_placebo
```
a. Shifting titers down by a full $\log_{10}$ unit pushes many vaccinees
   below the lowest titers observed among people like them, so the estimate
   relies on extrapolation (a near-violation of positivity) and is
   unstable, with a large standard error.
b. The capped shift answers a slightly different, more realistic
   question: the effect of raising titers by $\delta$ *for those who could
   plausibly reach the higher titer*, leaving others unchanged.
c. Since the placebo risk $p_0$ is estimated from the much larger placebo
   arm, its uncertainty is small; a fuller analysis would also propagate it
   (e.g., on the log scale).
:::
:::

## Wrap-up

- Both parts targeted causal parameters defined without a model — the ATE
  $\psi_b$ of smoking on each CpG, and the counterfactual risk $\psi_\delta$
  under a shift in antibody titers — and estimated them with machine
  learning and efficient influence functions.
- In both, a linear model gives an appealing shortcut: $\psi_b = \beta_1$,
  and $\psi_\delta - \psi_0 = \beta_1 \delta$. These identities hold *only*
  if the linear model is correct; when it is not, the coefficient answers a
  different (or impossible) question.
- Ideas from classical biomarker analysis still matter: variance moderation
  for high-dimensional testing (Part 1), and the sampling design (Part 2).
- In Part 1, the biomarker was the *outcome*. In Part 2, it was a
  *mediator* of the vaccine's effect. Decomposing a total effect into
  pathways through biomarkers — direct and indirect effects — is the subject
  of causal mediation analysis; see, e.g., the `medoutcon` and `medshift`
  packages (both on the course VM).
