# Part 2: Stochastic-interventional vaccine efficacy {#sec-sve}

```{r}
#| label: setup-sve
#| include: false
library(SuperLearner)
library(lmtp)
library(ggplot2)
source(here::here("practical/R/sim_vaccine_trial.R"))
source(here::here("practical/R/learners.R"))
set.seed(72104)
```

::: {.callout-note title="In a nutshell"}
We estimate how vaccine efficacy would change had the vaccine shifted
everyone's antibody titers. In our simulated trial, titers were measured only
in a case-cohort sample, so we use `lmtp` with weights for the sampling
design.

**Roadmap:** (1) the shift estimand, the SVE, and its linear-model analogue
→ (2) one shift, with misspecified and correct outcome GLMs and Super
Learner → (3) the SVE curve → (4) plug-in g-computation, and why it is more
fragile → (5) what happens if the sampling design is ignored → (6)
positivity.
:::

## The question

In vaccine trials, an *immune correlate of protection* is a marker of the
vaccine-induced immune response $S$ (e.g., neutralizing antibody titer)
that tracks protection against the outcome $Y$ (e.g., infection)
[@plotkin2012nomenclature]. Correlates help us understand *how* a vaccine
protects and can serve as surrogate endpoints for future trials
[@prentice1989surrogate; @gilbert2024four]. The question from the keynote:

> How would the risk of infection have changed had the vaccine induced
> higher, or lower, antibody titers?

The vaccine $A$ is randomized, but $S$ is not: it is a *post-treatment*
variable on the pathway $A \to S \to Y$, a mediator. Questions about
intervening on $S$ are questions about mediation.

## The data: a simulated trial with a case-cohort design

Data from the real trials are not public, so we simulate a trial of 10,000
people. The *complete* data on one person would be $X = (L, A, S, Y)$:

- $L$: standardized `age`, a `high_risk` behaviour indicator, `sex`;
- $A$: vaccine (1) or placebo (0), randomized 1:1;
- $S$: $\log_{10}$ neutralizing antibody titer ($S = 0$, the assay floor,
  in the placebo arm);
- $Y$: infection by the end of follow-up.

Assaying blood samples is expensive, so $S$ is measured only in a
*phase-two* sample, indicated by $R \in \{0, 1\}$: a case-cohort design
[@prentice1986case], a special case of two-phase sampling [@white1982two;
@breslow2003large]. The investigators choose who enters phase two, and they
choose using only $Y$ and $L$: every infected person (case) is sampled, and
non-cases are sampled at random within strata of `high_risk`. The sampling
probabilities are therefore known by design,
$$
\pi_0(Y, L) = \P_0(R = 1 \mid Y, L) =
\begin{cases}
1 & \text{if } Y = 1, \\
0.25 & \text{if } Y = 0 \text{ and } \texttt{high\_risk} = 1, \\
0.15 & \text{if } Y = 0 \text{ and } \texttt{high\_risk} = 0,
\end{cases}
$$
and recorded as `pi_R`. So the *observed* data unit is
$$
O = (L, A, R, RS, Y) \sim \P_0 \in \mathcal{P},
$$
where $RS$ is $S$ when $R = 1$ and missing otherwise.

```{tikz}
%| label: fig-dag-sve
%| echo: false
%| fig-cap: "DAG for the vaccine trial, in time order. The vaccine $A$ is randomized; covariates $L$ confound the effect of the immune marker $S$ on infection $Y$; inclusion $R$ in the phase-two sample depends, by design, only on $Y$ and $L$. (In our simulation, the vaccine acts on $Y$ only through $S$.)"
%| out-width: 70%
\usetikzlibrary{arrows.meta}
\tikzset{>={Stealth[length=2.2mm]}, every node/.style={minimum size=6mm},
  arc/.style={->, thick}}
\begin{tikzpicture}[node distance=2cm]
  \node (L) {$L$};
  \node[right of=L] (A) {$A$};
  \node[right of=A] (S) {$S$};
  \node[right of=S] (Y) {$Y$};
  \node[right of=Y] (R) {$R$};
  \draw[arc] (A) -- (S);
  \draw[arc] (S) -- (Y);
  \draw[arc] (Y) -- (R);
  % arcs from L enter later nodes at their tops, nested by length
  \draw[arc] (L.north) to[out=45, in=135] (S.north);
  \draw[arc] (L.north) to[out=55, in=125] (Y.north);
  \draw[arc] (L.north) to[out=65, in=115] (R.north);
\end{tikzpicture}
```

```{r}
#| label: sim-data
trial <- sim_vaccine_trial()
head(trial)
with(trial, table(vaccine = A, infected = Y))
with(trial, table(vaccine = A, phase_two = R))
with(trial, table(phase_two = R, infected = Y))  # every case is sampled
```

We focus on vaccine recipients in the phase-two sample:

```{r}
#| label: vax-data
vax <- subset(trial, A == 1 & R == 1)
nrow(vax)
mean(vax$Y)   # cases are heavily over-represented!
```

## The target parameter

Let $\overline{Q}_0(s, a, l) = \E_0[Y \mid S = s, A = a, L = l]$ be the risk
of infection, and $g_{0,S}(s \mid a, l)$ the conditional density of $S$ given
$A = a$ and $L = l$ (the *generalized propensity score*). A *shift
intervention* replaces each vaccinee's titer $S$ with $d(S, L) = S + \delta$
[@diaz2012population; @haneuse2013estimation]. Our target parameter is the
**stochastic-interventional vaccine efficacy** [@hejazi2020efficient;
@gilbert2021covpn],
$$
\psi_{0,\delta} = 1 - \frac{\theta_{0,\delta}}{\rho_0}.
$$

- The **numerator** $\theta_{0,\delta}$ is the risk of infection had the
  vaccine been given ($A = 1$) and each vaccinee's titer been shifted by
  $\delta$. Under no unmeasured confounding of $S$ and $Y$ given $L$, and
  positivity, it is identified by
  $$
  \theta_{0,\delta} = \E_0 \left[ \int \overline{Q}_0(s + \delta, 1, L) \,
  g_{0,S}(s \mid 1, L) \, ds \right]:
  $$
  draw a titer from its distribution among vaccinees with covariates $L$,
  shift it by $\delta$, and average the vaccinee risk at the shifted titer
  over $L$.
- The **denominator** $\rho_0$ is the risk of infection had everyone
  received placebo ($A = 0$). Since $A$ is randomized, it is identified by
  the placebo-arm risk, $\rho_0 = \P_0(Y = 1 \mid A = 0)$.^[Randomization
  also gives the covariate-adjusted identification $\rho_0 = \E_0[\P_0(Y = 1
  \mid A = 0, L)]$. Estimating $\rho_0$ this way uses the prognostic
  information in $L$ and can improve efficiency; we use the simpler
  unadjusted estimate.]

At $\delta = 0$, this is the usual vaccine efficacy; as $\delta$ varies, it
traces out how efficacy would change with the vaccine's immunogenicity. It was
one of the immune correlates analyses of the COVE trial of the Moderna
mRNA-1273 COVID-19 vaccine [@gilbert2021covpn-cove; @hejazi2023stochastic;
@huang2023stochastic]. Unlike the controlled vaccine efficacy from the keynote
[@gilbert2022controlled], which sets every vaccinee's titer to the *same*
value $s$, a shift moves each person's titer from where it naturally fell.

### Exercise 2.1 ✏️ {#sec-ex21}

::: {.callout-note title="Shifts, linear models, and sampling"}
a. Suppose the outcome model is linear: $\overline{Q}_0(s, l) = \beta_0 +
   \beta_1 s + \gamma^\top l$. Show that $\theta_{0,\delta} - \theta_{0,0} =
   \beta_1 \delta$. What does $\psi_{0,\delta}$ look like as a function of
   $\delta$?
b. What goes wrong with the linear model as $\delta$ grows? (Hint:
   $\theta_{0,\delta}$ is a probability.)
c. Among the phase-two vaccinees, what is the share of cases, compared to
   the full vaccine arm? What would happen if we estimated $\theta_{0,\delta}$
   from `vax` as if it were a random sample of vaccinees? How do the
   weights $1 / \pi_0(Y, L)$ fix this?
d. Positivity for a shift intervention requires that, for each $l$, the
   shifted value $s + \delta$ is a value of $S$ that could occur among
   people with $L = l$. Why is this harder to satisfy for large $|\delta|$?
:::

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
a. $\theta_{0,\delta} = \E_0[\beta_0 + \beta_1 (S + \delta) + \gamma^\top
   L \mid A = 1] = \theta_{0,0} + \beta_1 \delta$. So $\psi_{0,\delta} = 1 -
   (\theta_{0,0} + \beta_1 \delta) / \rho_0$ is a straight line in $\delta$
   with slope $-\beta_1 / \rho_0$: the linear
   model's $\beta_1$ *is* the effect of a unit shift, just as $\beta_1$ was
   the ATE in Part 1.
b. A straight line eventually leaves $[0, 1]$: for large positive $\delta$
   (with $\beta_1 < 0$), the predicted risk becomes negative and the SVE
   exceeds 100%; the model also forces each additional unit of titer to
   buy the same reduction in risk, whatever the starting titer.
c. All cases are sampled but only 15–25% of non-cases, so cases are hugely
   over-represented (compare `mean(vax$Y)` with the infection rate in the
   vaccine arm). Ignoring this would badly overestimate risk. Weighting each
   phase-two person by $1 / \pi_0(Y, L)$ makes them stand in for the
   $1 / \pi_0(Y, L)$ people like them in the full trial (inverse probability
   weighting), recovering the full-cohort target.
d. Shifting by $\delta$ asks about titers $s + \delta$; for large
   $|\delta|$, many shifted titers fall outside the range observed among
   similar people (e.g., below the assay floor, or far above any observed
   response), so the data contain no information about risk there and any
   estimate is an extrapolation.
:::
:::

## Estimation with `lmtp`

The `lmtp` package [@diaz2023nonparametric] implements estimators of the
effects of *modified treatment policies*, of which our shift is one.
`lmtp_tmle()` estimates the numerator $\theta_{0,\delta}$ by targeted
maximum likelihood. It needs two nuisance estimates: $\overline{Q}_n$, and
an estimate of the density ratio $g_{0,S}(s - \delta \mid 1, l) / g_{0,S}(s
\mid 1, l)$, which `lmtp` obtains by classifying observed versus shifted
titers. Both are fit with `SuperLearner` libraries (`learners_outcome`,
`learners_trt`). `lmtp_tmle()` accepts sampling weights through `weights`,
which gives the inverse-probability-weighted version of the estimator for
two-phase designs [@rose2011targeted].^[The
estimator in the `txshift` package [@hejazi2020txshift] also *augments* the
weighted estimator with a model for the sampling mechanism, using information
on the whole cohort to gain efficiency [@breslow2009using;
@hejazi2020efficient]. The reweighted version is valid here because $\pi_0$
is known by design.]

In our simulation, the risk of infection falls with titer only up to a
threshold ($\log_{10}$ titer of 1.2) and is flat above it. A main-terms
logistic regression (`"SL.glm"`) for $\overline{Q}_0$ is therefore
misspecified; `"SL.glm_threshold"` (in `practical/R/learners.R`), which uses
$\min(S, 1.2)$ in place of $S$, is correctly specified. For the density
ratio, we use a main-terms logistic regression throughout (see
Exercise 2.6c).

### Exercise 2.2 💻 {#sec-ex22}

::: {.callout-note title="One shift, three outcome models"}
Estimate $\theta_{0,0.5}$, the risk among vaccinees had every vaccinee's titer
been $0.5$ $\log_{10}$ units (about 3-fold) higher, using for
$\overline{Q}_n$ (i) the misspecified GLM, (ii) the correctly specified GLM,
and (iii) a Super Learner whose library includes the correct GLM. Compare
to the truth, `true_sve(0.5)`.
:::

```{r}
#| label: ex22-skeleton
#| eval: false
shift_by <- function(delta) function(data, trt) ___
fit_shift <- function(delta, q_lib, weights = 1 / vax$pi_R) {
  lmtp_tmle(
    data = vax, trt = "S", outcome = "Y",
    baseline = c("age", "high_risk", "sex"),
    shift = shift_by(delta), outcome_type = "binomial",
    learners_outcome = q_lib, learners_trt = "SL.glm", folds = 5,
    weights = weights
  )
}
sl_lib <- c("SL.mean", "SL.glm", "SL.glm_threshold", "SL.glmnet",
            "SL.earth", "SL.gam")
fits_05 <- list(glm_wrong = fit_shift(0.5, "SL.glm"),
                glm_right = fit_shift(0.5, ___),
                super_learner = fit_shift(0.5, ___))
sapply(fits_05, function(fit) c(theta = fit$estimate@x,
                                  se = fit$estimate@std_error))
true_sve(0.5)
```

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| label: ex22-solution
shift_by <- function(delta) function(data, trt) data[[trt]] + delta
fit_shift <- function(delta, q_lib, weights = 1 / vax$pi_R) {
  lmtp_tmle(
    data = vax, trt = "S", outcome = "Y",
    baseline = c("age", "high_risk", "sex"),
    shift = shift_by(delta), outcome_type = "binomial",
    learners_outcome = q_lib, learners_trt = "SL.glm", folds = 5,
    weights = weights
  )
}
sl_lib <- c("SL.mean", "SL.glm", "SL.glm_threshold", "SL.glmnet",
            "SL.earth", "SL.gam")
fits_05 <- list(glm_wrong = fit_shift(0.5, "SL.glm"),
                glm_right = fit_shift(0.5, "SL.glm_threshold"),
                super_learner = fit_shift(0.5, sl_lib))
sapply(fits_05, function(fit) c(theta = fit$estimate@x,
                                  se = fit$estimate@std_error))
true_sve(0.5)
```
All three are close to the truth, *including* the misspecified GLM. The
TML estimator is doubly robust: it is consistent if either $\overline{Q}_n$
or the density ratio is consistent, and here the density ratio model is
correctly specified (Exercise 2.6c). Exercise 2.4 shows what happens without
that protection.
:::
:::

### Exercise 2.3 💻 {#sec-ex23}

::: {.callout-note title="The SVE curve"}
Estimate $\theta_{0,\delta}$ over a grid of shifts with the Super Learner,
compute $\psi_{n,\delta}$ using the placebo-arm risk $\rho_n$ (estimable from
the full trial, since $S$ is not needed), and compare to the truth, which we
know because we simulated the data (`true_sve()`).
:::

```{r}
#| label: ex23-skeleton
#| eval: false
deltas <- c(-0.5, -0.25, 0, 0.25, 0.5, 0.75)
risk_placebo <- mean(trial$Y[trial$A == ___])
theta_n <- sapply(deltas, function(delta) fit_shift(___)$estimate@x)
sve <- data.frame(delta = deltas, lmtp = 1 - theta_n / ___,
                  truth = true_sve(deltas)$sve)
sve
```

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| label: ex23-solution
deltas <- c(-0.5, -0.25, 0, 0.25, 0.5, 0.75)
risk_placebo <- mean(trial$Y[trial$A == 0])
theta_n <- sapply(deltas, function(delta) fit_shift(delta, sl_lib)$estimate@x)
sve <- data.frame(delta = deltas, lmtp = 1 - theta_n / risk_placebo,
                  truth = true_sve(deltas)$sve)
sve
```
The estimates track the truth for small shifts and drift upward for the
larger positive ones, where more vaccinees are shifted beyond the titers
observed among people like them. This is a first sign of the positivity
strain explored in Exercise 2.6a.
:::
:::

### Exercise 2.4 💻 {#sec-ex24}

::: {.callout-note title="The linear-model analogue"}
Now estimate the same curve by *g-computation*: plug an outcome regression
into the identification formula, with no correction. Use parametric models fit by weighted regression to `vax`: (i) the linear
probability model of Exercise 2.1(a), (ii) the misspecified main-terms
logistic regression, and (iii) the correctly specified threshold logistic
regression. Predict each vaccinee's risk at $S + \delta$, average (with the
weights), and convert to SVE. Plot all the curves.
:::

```{r}
#| label: ex24-skeleton
#| eval: false
w <- 1 / vax$pi_R
fit_lin <- lm(Y ~ S + age + high_risk + sex, data = vax, weights = w)
fit_logit <- glm(Y ~ S + age + high_risk + sex, data = vax, weights = w,
                 family = quasibinomial())
fit_thresh <- glm(Y ~ pmin(S, 1.2) + age + high_risk + sex, data = vax,
                  weights = w, family = quasibinomial())
gcomp_sve <- function(fit, delta) {
  risk <- predict(fit, newdata = transform(vax, S = ___), type = "response")
  1 - weighted.mean(risk, w) / risk_placebo
}
sve$linear <- sapply(deltas, gcomp_sve, fit = fit_lin)
sve$logistic <- sapply(deltas, gcomp_sve, fit = fit_logit)
sve$threshold <- sapply(deltas, gcomp_sve, fit = fit_thresh)
```

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| label: ex24-solution
w <- 1 / vax$pi_R
fit_lin <- lm(Y ~ S + age + high_risk + sex, data = vax, weights = w)
fit_logit <- glm(Y ~ S + age + high_risk + sex, data = vax, weights = w,
                 family = quasibinomial())
fit_thresh <- glm(Y ~ pmin(S, 1.2) + age + high_risk + sex, data = vax,
                  weights = w, family = quasibinomial())
gcomp_sve <- function(fit, delta) {
  risk <- predict(fit, newdata = transform(vax, S = S + delta),
                  type = "response")
  1 - weighted.mean(risk, w) / risk_placebo
}
sve$linear <- sapply(deltas, gcomp_sve, fit = fit_lin)
sve$logistic <- sapply(deltas, gcomp_sve, fit = fit_logit)
sve$threshold <- sapply(deltas, gcomp_sve, fit = fit_thresh)
round(sve, 3)

# the linear model's slope: psi(delta) = psi(0) - beta_1 * delta / rho
-coef(fit_lin)[["S"]] / risk_placebo
```

```{r}
#| label: ex24-plot
#| fig-cap: "SVE as a function of the shift in log10 titer."
sve_long <- reshape(sve, direction = "long", varying = names(sve)[-1],
                    v.names = "sve", timevar = "method",
                    times = names(sve)[-1], idvar = "delta")
ggplot(sve_long, aes(delta, sve, colour = method)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 1, linetype = "dotted") +
  labs(x = expression(paste("Shift ", delta, " in ", log[10], " titer")),
       y = "SVE", colour = NULL) +
  theme_bw()
```

The linear model forces a straight line and soon claims an efficacy above
100%: its "risk" is negative. The misspecified logistic regression keeps
predicting benefit from higher titers where there is none, overstating the
SVE for positive shifts; the threshold model tracks the truth. Plug-in
g-computation is only as good as its outcome model. In contrast, `lmtp` was
protected by its density ratio model in Exercise 2.2, even with the
misspecified GLM.
:::
:::

### Exercise 2.5 💻 {#sec-ex25}

::: {.callout-note title="Forgetting the design"}
Re-estimate $\psi_{0,0.5}$ with `lmtp_tmle()` and the Super Learner,
**without** the sampling weights. What happens, and why?
:::

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| label: ex25-solution
fit_unwt <- fit_shift(0.5, sl_lib, weights = NULL)
c(weighted = 1 - fits_05$super_learner$estimate@x / risk_placebo,
  unweighted = 1 - fit_unwt$estimate@x / risk_placebo,
  truth = true_sve(0.5)$sve)
```
Without weights, the phase-two sample is treated as representative of all
vaccinees, but cases are over-represented about five-fold, so the estimated
risk is far too high and the SVE far too low. The estimator must account for the
two-phase design.
:::
:::

### Exercise 2.6 *(stretch)* {#sec-ex26}

::: {.callout-note title="Positivity, realistic shifts, and the density ratio"}
a. 💻 Estimate $\psi_{0,1}$, and compare to `true_sve(1)`. Repeat with
   `"SL.glm_threshold"` as the outcome model, and compare to its
   g-computation counterpart from Exercise 2.4. What might explain the
   result? (Hint: what is the range of `vax$S`?)
b. 💻 A shift that respects positivity only shifts people whose shifted
   titer remains plausible [@hejazi2020efficient]: $d(s, l) = s + \delta$ if
   $s + \delta \leq u$, and $d(s, l) = s$ otherwise. Implement it with $u$
   the 90th percentile of $S$ in the phase-two sample, and re-estimate the
   SVE curve. How would you interpret this effect?
c. ✏️ In the simulation, $S \mid A = 1, L$ is normal with mean $m(L)$
   linear in $L$ and constant variance $\sigma^2$. Show that the log density
   ratio, $\log\{g_{0,S}(s - \delta \mid 1, l) / g_{0,S}(s \mid 1, l)\}$,
   is linear in $(s, l)$, so that the main-terms logistic classifier for the
   density ratio is correctly specified.
:::

::: {.content-hidden when-profile="students"}
::: {.callout-tip collapse="true" title="Solution"}
```{r}
#| label: ex26-solution
# (a) a large shift
range(vax$S)
c(sl = 1 - fit_shift(1, sl_lib)$estimate@x / risk_placebo,
  glm_right = 1 - fit_shift(1, "SL.glm_threshold")$estimate@x / risk_placebo,
  gcomp_right = gcomp_sve(fit_thresh, 1),
  truth = true_sve(1)$sve)

# (b) a positivity-respecting shift, capped at the 90th percentile
u <- quantile(vax$S, 0.9)
shift_capped <- function(delta) {
  function(data, trt) ifelse(data[[trt]] + delta <= u, data[[trt]] + delta,
                             data[[trt]])
}
psi_capped <- sapply(deltas, function(delta) {
  lmtp_tmle(
    data = vax, trt = "S", outcome = "Y",
    baseline = c("age", "high_risk", "sex"),
    shift = shift_capped(delta), outcome_type = "binomial",
    learners_outcome = sl_lib, learners_trt = "SL.glm", folds = 5,
    weights = 1 / vax$pi_R
  )$estimate@x
})
round(data.frame(delta = deltas, capped = 1 - psi_capped / risk_placebo,
                 uncapped = sve$lmtp), 3)
```
a. Shifting titers up by a full $\log_{10}$ unit (10-fold) asks about titers
   well beyond those observed for many vaccinees. The density ratio becomes
   extreme and the TML estimate less stable, even with a correct outcome
   model. Plug-in g-computation with the correct model, by contrast, simply
   extrapolates its (correct) threshold. This is a near-violation of
   positivity: the data say little about such large shifts.
b. The capped shift answers a different, more realistic question: the
   effect of raising titers by $\delta$ *for those who could plausibly reach
   the higher titer*, leaving others unchanged. Its efficacy gains are
   smaller, because fewer people are shifted.
c. With $g_{0,S}(s \mid 1, l) \propto \exp\{-(s - m(l))^2 / (2\sigma^2)\}$,
   $$
   \log \frac{g_{0,S}(s - \delta \mid 1, l)}{g_{0,S}(s \mid 1, l)}
   = \frac{(s - m(l))^2 - (s - \delta - m(l))^2}{2 \sigma^2}
   = \frac{\delta}{\sigma^2} \Big(s - m(l) - \frac{\delta}{2}\Big),
   $$
   which is linear in $s$ and (since $m$ is linear) in $l$. The classifier
   of shifted versus observed titers estimates exactly this log-odds, so a
   main-terms logistic regression is correctly specified.
:::
:::

## Wrap-up

- Both parts targeted causal parameters defined without a model: the ATE
  $\psi_{0,b}$ of smoking on each CpG, and the SVE $\psi_{0,\delta}$ under
  a shift in antibody titers. We estimated both with machine learning and
  efficient influence functions.
- In both, a linear model gives an appealing shortcut: $\psi_{0,b} =
  \beta_1$, and $\theta_{0,\delta} - \theta_{0,0} = \beta_1 \delta$. These
  identities hold *only* if the linear model is correct; when it is not, the
  coefficient answers a different (or impossible) question.
- A misspecified GLM breaks a plug-in estimator, but not a doubly robust one
  whose other nuisance is right. Given a library that contains a good model,
  the Super Learner finds it.
- Ideas from classical biomarker analysis still matter: variance moderation
  for high-dimensional testing (Part 1), and the sampling design (Part 2).
- In Part 1, the biomarker was the *outcome*. In Part 2, it was a
  *mediator* of the vaccine's effect. Decomposing a total effect into
  pathways through biomarkers (direct and indirect effects) is the subject
  of causal mediation analysis; see, e.g., the `medoutcon` and `medshift`
  packages (both on the course VM).
